Tools for Identifying Partisan Gerrymandering (PA)

### 2019-06-14 Latex/_maintext.tex
main update

        \footnote{There is a debate about how to count split counties, which was evident when different parties in the case counted different total county splits for the same plan. One way to count splits to count the total number of counties that are in at least two districts. The other way is to count the number of times a county is split, i.e., the number of separate pieces into which the county is divided. In the first counting method, a county can have portions of itself found in 18 different districts, but have this counted as only one county split. Under the second method, this wholesale splitting of the county would be counted as 18 splits. While both methods provide useful information, we regard the second as more informative and it is the one reported in Table 1. However, we must also be alert for divisions of a county into two or more discontiguous pieces found within the same district. We call this “fracking”, to provide a term similar sounding to ‘packing’ and ‘cracking.’ We find “fracking” a particularly egregious problem since there is no need for counties to be fracked, and fracking may be used as a gerrymandering tool. Instances where counties have multiple discontiguous pieces located in a single district are not discussed in the Pennsylvania Supreme Court ruling on the unconstitutionality of the 2011 Pennsylvania congressional plan. However, our analysis of the 2011 plan reveals several cases of fracking: two of the most egregious are districts 14 and 18 in Allegheny County (Pittsburgh). This concept has also been addressed by the U.S. District Court in North Carolina (\textit{Common Cause v. Rucho} No 1:16-CV-1026 (U.S. District Court, Middle District of North Carolina, 2018, slip op at p. 105 [p. 194])).}

@@

    Each of the three measures in Table \ref{tab:summaries} indicates a very dramatic deviation from good government standards in the 2011 Enacted map. And, each of three metrics order the plans in good government terms from worst to best in exactly the same way, with the worst plan being the 2011 Enacted map, the next worst being the Republican remedy plan, followed by the the Democratic remedy plan, and the very best being the court-ordered map. Moreover, as we will see later, the same ordering of plans is reflected when we examine the partisan implications of the various maps, with the map ordered into place as a remedy by the Pennsylvania Supreme Court close to perfectly neutral in its expected partisan effects once we recognize the implications of the partisan geography in Pennsylvania in ``naturally” tilting outcomes slightly in a pro-Republican direction because of differences in the degree of geographic concentration of each party’s supporters. (see below) Moreover, we show that the proposed Republican remedy plan can be characterized as a ``disguised” pro-Republican partisan gerrymander -- what we call a ``stealth gerrymander” -- which is nearly as pernicious in its expected pro-Republican partisan consequences as the 2011 plan whose effects it claims to ``remedy".

@@

        \footnote{For consistency, we report the average vote share after imputing 0.25 and 0.75 for uncontested districts. In raw votes, Democrats won more votes than Republicans in 2012, but by taking the imputed average, it shows they won a minority of votes due to disproportionate uncontested districts. The Republican raw vote shares are 49.2\% in 2012, 55.5\% in 2014, and 54.1\% in 2016.}

@@

@@

% SUBSECTION

@@

    There are other indicators of statewide partisan gerrymandering effects that also provide a compelling picture of the 2011 map as an (enduring) partisan gerrymander. There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) \textit{declination} \citep{Warrington2018}. Each of these measures can be directly calculated for the 2011 map based on district election outcomes. 

@@

@@

    \textit{Partisan Bias} indicates asymmetry in the translation of each party’s votes into seats. Customarily, partisan bias is measured at the (hypothetical) point where each party’s share of the two-party vote is exactly 50\%. At that point, both parties should, by symmetry, receive identical seats shares. Thus, the difference between a party’s projected seat-share with 50\% of the vote and exactly 50\% can be taken as a measure of partisan bias. Note, however, that although a plan must be proportional at the ($50,50$) point, a partisan bias of zero does not imply proportional representational. What partisan bias taps is differential treatment of the parties? Thus, for example, if party R gets 56\% of the seats with 52\% of the vote, this is not a sign of bias as long as party L also can be expected to get 56\% of the seats were it to have 52\% of the vote.
        \footnote{Although the basic idea goes back at least as far as \citet{Dahl1956}, the modern \textit{locus classicus} in political science for the idea of partisan bias(and the parallel concept of responsiveness/swing ratio) is \citet{Tufte1973}, with important recent work done by Gary King (see e.g., \citep{GelmanKing1994_unifiedAJPS}; and review in \citet{GrofmanKing2007_ELJ}). A closely related line of research is found in the political geography literature (with \textit{locus classicus} \citet{Brookes1959, Brookes1960} -- with some further methodological improvements by scholars such as \citet{Johnston1994, Rossiter1997} and \citet{Johnston2002}.}   
    There are a number of different ways to estimate partisan bias based on the shape of the votes to seats distribution \citep{Tufte1973, Grofman1983, Browning_King_1987_seats_votes, GelmanKing1994_unifiedAJPS, Grofman_et_al_1997_SwingRatio_Bias, Zingher2016_bias_swingratio_JEPP}. The method we use here is basically that of \citet{Tufte1973} with some additional complexity building in ideas of \citet{GelmanKing1994_unifiedAJPS} on simulating potential outcomes in a probabilistic fashion. We adapted the code from \citet{Kastellec_et_al_2008_PS}, which parallels that used by \texttt{JudgeIt}. It models inter-election shifts and adjusts for differences over time in vote shares among and between districts in a probabilistic fashion, using a simulation with 1000 repetitions (see more below). We then apply the following procedure to find the mean partisan bias and range of expected bias over the course of the decade.
    Hypothetical elections are constructed by incrementally adding (or subtracting) one percentage point (on average) at a time to find aggregate seat outcomes under differing mean vote shares. The resulting ($s$, $v$) points on the votes-to-seats curve are then converted to a log odds form by using $ log(\frac{s}{1-s}) $ as the dependent variable, and $ log(\frac{v}{1-v}) $ as the independent variable. A regression on the transformed variables is then run, with transformed $s$ as the dependent variable, for $v$ values in some range near a 50\% vote share. 

@@

    \textit{Partisan Bias} is then calculated from the intercept of this regression using an exponential transformation (for details see \citet{Grofman1983}). Partisan bias is calculated from the Republican perspective, and \underline{negative} values indicates bias against Republicans.

@@

    Among the more recent measures of gerrymandering is the \textit{Declination}. \textit{Declination} is an astronomical term referring to the angle on a compass, with a Latin root meaning ``a bending away". For gerrymandering, it uses the angles created from ordering all of the districts by vote share, computing the mean vote-share for each party separately for the seats they won, and then comparing the differences in their distribution. To compare, a line is draw from the mean vote-share to the 0.5 line separately for each party. The angle formed from the different slopes of the line is called the \textit{declination}. An angle indicates whether a plan is a gerrymandering, under the logic that vote-shares should be distributed in some uniform manner across districts. If partisan are \textit{packed}, the angle will be greater. All values range between -1 and 1. Though we have defined the vote-share in terms of Republican outcomes, we adjusted the values such that \underline{negative} values are favorable to the Democrats. Further information can be found in \citet{Warrington2018}.  

>>>>

        \footnote{There is a debate about how to count split counties, which was evident when different parties in the case counted different total county splits for the same plan. One way to count splits to count the total number of counties that are in at least two districts. The other way is to count the number of times a county is split, i.e., the number of separate pieces into which the county is divided. In the first counting method, a county can have portions of itself found in 18 different districts, but have this counted as only one county split. Under the second method, this wholesale splitting of the county would be counted as 18 splits. While both methods provide useful information, we regard the second as more informative and it is the one reported in Table 1. However, we must also be alert for divisions of a county into two or more discontiguous pieces found within the same district. We call this ``fracking”, to provide a term similar sounding to ‘packing’ and ‘cracking.’ We find fracking a particularly egregious problem since there is no need for counties to be fracked, and fracking may be used as a gerrymandering tool. Instances where counties have multiple discontiguous pieces located in a single district are not discussed in the Pennsylvania Supreme Court ruling on the unconstitutionality of the 2011 Pennsylvania congressional plan. However, our analysis of the 2011 plan reveals several cases of fracking: two of the most egregious are districts 14 and 18 in Allegheny County (Pittsburgh). This concept has also been addressed by the U.S. District Court in North Carolina (\textit{Common Cause v. Rucho} No 1:16-CV-1026 (U.S. District Court, Middle District of North Carolina, 2018, slip op at p. 105 [p. 194])).}

@@

    Each of the three measures in Table \ref{tab:summaries} indicates a very dramatic deviation from good government standards in the 2011 Enacted map. And, each of three metrics order the plans in good government terms from worst to best in exactly the same way, with the worst plan being the 2011 Enacted map, the next worst being the Republican remedy plan, followed by the the Democratic remedy plan, and the very best being the court-ordered map. Moreover, as we will see later, the same ordering of plans is reflected when we examine the partisan implications of the various maps, with the map ordered into place as a remedy by the Pennsylvania Supreme Court close to perfectly neutral in its expected partisan effects once we recognize the implications of the partisan geography in Pennsylvania in ``naturally” tilting outcomes slightly in a pro-Republican direction because of differences in the degree of geographic concentration of each party’s supporters (see below). Moreover, we show that the proposed Republican remedy plan can be characterized as a ``disguised” pro-Republican partisan gerrymander -- what we call a ``stealth gerrymander” -- which is nearly as pernicious in its expected pro-Republican partisan consequences as the 2011 plan whose effects it claims to ``remedy".

@@

        \footnote{For consistency, we report the average vote share after imputing 0.25 and 0.75 for uncontested districts. In raw votes, Democrats won more votes than Republicans in 2012, and the difference is an artifact of the plan's `packing' and `cracking'. The Democratic raw vote shares are 50.8\% in 2012, 44.5\% in 2014, and 45.9\% in 2016.}

@@

    There are other indicators of statewide partisan gerrymandering effects that also provide a compelling picture of the 2011 map as an (enduring) partisan gerrymander. There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) \textit{declination} \citep{Warrington2018} (details below). Each of these measures can be directly calculated for the 2011 map based on district election outcomes in 2012, 2014, and 2016. 

@@

	To estimate these metrics for plans that have no electoral data associated with them, we add some additional complexity building on the ideas of \citet{GelmanKing1994_unifiedAJPS} on simulating potential outcomes in a probabilistic fashion. We adapted the code from \citet{Kastellec_et_al_2008_PS}, which parallels that used by \texttt{JudgeIt}. It models inter-election shifts and adjusts for differences over time in vote shares among and between districts in a probabilistic fashion, using a simulation with 1,000 repetitions (see more below). We shift the districts to approximate each party recieving exactly 50\% of the state-wide vote. The same procedure is applied to the 2011 enacted map so direct comparisons between the plans can be made.

@@

    \textit{Partisan Bias} indicates asymmetry in the translation of each party’s votes into seats. Customarily, partisan bias is measured at the (hypothetical) point where each party’s share of the two-party vote is exactly 50\%. At that point, both parties should, by symmetry, receive identical seats shares. Thus, the difference between a party’s projected seat-share with 50\% of the vote and exactly 50\% can be taken as a measure of partisan bias. Note, however, that although a plan must be proportional at the ($50,50$) point, a partisan bias of zero \underline{does not} imply proportional representational. What partisan bias taps is the differential treatment of parties. Thus, for example, if party R gets 61\% of the seats with 52\% of the vote, this is not a sign of bias as long as party L also can be expected to get 61\% of the seats were it to have 52\% of the vote.

@@

        \footnote{Although the basic idea goes back at least as far as \citet{Dahl1956}, the modern \textit{locus classicus} in political science for the idea of partisan bias (and the parallel concept of responsiveness/swing ratio) is \citet{Tufte1973}, with important recent work done by Gary King (see e.g., \citep{GelmanKing1994_unifiedAJPS}; and reviewed in \citet{GrofmanKing2007_ELJ}). A closely related line of research is found in the political geography literature (with \textit{locus classicus} \citet{Brookes1959, Brookes1960} -- with some further methodological improvements by scholars such as \citet{Johnston1994, Rossiter1997} and \citet{Johnston2002}.}   

@@

    There are a number of different ways to estimate partisan bias based on the shape of the votes to seats distribution \citep{Tufte1973, Grofman1983, Browning_King_1987_seats_votes, GelmanKing1994_unifiedAJPS, Grofman_et_al_1997_SwingRatio_Bias, Zingher2016_bias_swingratio_JEPP}. The method we use here is that of \citet{Tufte1973}. Hypothetical elections are constructed by incrementally adding (or subtracting) one percentage point (on average) at a time to find aggregate seat outcomes under differing mean vote shares. The resulting ($s$, $v$) points on the votes-to-seats curve are then converted to a log odds form by using $ log(\frac{s}{1-s}) $ as the dependent variable, and $ log(\frac{v}{1-v}) $ as the independent variable. A regression on the transformed variables is then run, with transformed $s$ as the dependent variable, for $v$ values in some range near a 50\% vote share. 

@@

    \textit{Partisan Bias} is then calculated from the intercept of this regression using an exponential transformation (for details see \citet{Grofman1983}). Partisan bias is calculated from the Republican perspective, and \underline{negative} values indicates bias against Republicans. We do this for each of the 1,000 simulated elections for each of the maps, and take the average.

@@

    Among the more recent measures of gerrymandering is the \textit{Declination}. \textit{Declination} is an astronomical term referring to the angle on a compass, with a Latin root meaning ``a bending away". For gerrymandering, it uses the angles created from ordering all of the districts by vote share, computing the mean vote-share for each party separately for the seats they won, and then comparing the differences in their distribution. To compare, a line is draw from the mean vote-share to the 0.5 line separately for each party. The angle formed from the different slopes of the line is called the \textit{declination}. An angle indicates whether a plan is a gerrymandering, under the logic that vote-shares should be distributed in some uniform manner across districts. If partisan are \textit{packed}, the angle will be greater. All values range between -1 and 1. Though we have defined the vote-share in terms of Republican outcomes, we adjusted the values such that \underline{negative} values are favorable to the Democrats. Further information on declination can be found in \citet{Warrington2018}.  
